test_imputed

### 2015-11-09 finalwriteup/draft.rmd
save early nov

Project decision makers have struggled with forecasting project costs accurately for decades, across many industries. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% (Flyvbjerg, 2006). A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget on average (Flyvbjerg and Budzier, 2011). These statistics cast a concerning glimpse at the financial story playing out behind the scenes of these projects which can cause job losses, ruin private businesses, and burn through government balancesheets. Complex long term projects are necessary to produce  railways, houses, websites, and movies. Depending on the industry, the the financial risk of completing a project on budget is transferred to consulting companies who are enlisted to complete portions of a project. The risk is transferred by requiring a fixed flat fee from the consultant, meaning they commit to complete their role for a nominated price. In a similar way to infrastructure and IT projects, consultants are susceptible to underestimating the amount of their time that will be spent in their role, and must wear the cost of extra employee hours spent. For privately owned businesses, mitigating this risk would increase the chances of the business surviving as well as growing and succeeding.

The ramifications of unprofitable projects in a private business are multifold. Employee morale deteoriates and employees may view themselves as incompetent if they are responsible for the project's delivery. Employees also experience stress attempting to complete the project within a disappearing budget and may perform lower quality work in exchange for speed. This is unfortunate if in reality, the project fee was under estimated. Unprofitable projects also limit a business' opportunity to invest money into marketing or staff training, which would improve a business' standing. The financial risk taken on fixed flat fee projects may also discourage business' from taking different risks with better potential pay-offs, such as expanding the business into a new area. Clearly, predicting fixed fee project costs more accurately could positively affect a business' growth opportunities and staff morale.

@@

The challenge for consultants to reduce the risk of a project going over budget, in terms of hours spent, is not easily solved. The reason is that complex projects are always different from each other, and similar work may take significantly different amounts of time. For example, an ad filmed in a remote location for chocolate will cost significantly more than an ad filmed on an exisitng set, even though the product structure is almost identical. Therefore, a fixed product price, which you could apply to products such as appliances, is inappropriate for film projects. One option is charging for consultants' time and the hiring of materials by the hour; a zero risk fee structure. However in many industries this is not competitive enough. Other consultancies may be willing to offer a fixed flat fee for the same work, reducing financial risk for the client, and luring the client away. Another fee strategy may be to track worker hours, and stop work once the consulting fee has been depleted. Then, negotiate further fees, or variations, with the client and persuade them to accept. This also satisfactorily reduces consultant financial risk, and in some cases, depending on the nature of the work or uncontrollable changes, this method can be effective. However, if the project has been following a predictable process but the hours are nevertheless inflating beyond expectation (a common situation), consultants may prefer to wear losses or low profits to preserve the relationship with their client. Consultant's may believe it is in their best interest to maintain a reliable, trustworthy reputation in their industry over creating financial friction. Clearly, there is no straightforward solution to reducing the financial risk taken on by consultant's who offer fixed fees in complex projects. In some indusries, fixed flat fee structures are expected and business' must conform to survive.

@@

Limited research is available on the cost estimation techniques performed by industry for complex projects, however two surveys were performed in 2000: one on construction projects, the other on IT projects. The construction cost estimation study by @Akintoye2000 surveyed 84 UK construction contractors, ranging from small to medium to large, about their estimating practises. They found that the main method for cost estimation was breaking the project into detailed parts and adding up the cost of each item. The next two most popular methods were comparison with similar projects based on documented facts, and comparison with similar projects based on personal experience. These can all be classified as experience based models [@Akintoye2000]. A survey by Moores and Edwards in 1992 of 54 software developing companies found that detailed project planning tools were used by most companies as opposed to cost estimation tools, suggesting that projects were priced based on an analysis of a detailed breakdown of tasks within a project. In the same survey, 91% of software companies cited cost estimation as a problem. Although limited recent literature is available on industry cost estimation practises, historically it can be observed that for complex construction and software projects, cost predictions are chiefly made with a mixture of experience, project comparisons, and an analysis of the project details and tasks. It is also acknowledged that cost estimation is a problem and hypothesised that inexperience and lack of time are chief contributors [@Akintoye2000].

@@

@Lovallo2003 tackled the psychology behind the high failure rates of executives in predicting costs of projects such as manufacturing plant construction, mergers and acquisitions, large infrastructure and software development. Their theory stems from Kahneman's work on decision making that won him the nobel prize for economics in 2002. The research argues that humans' natural optimistic view of their own skills leads to consistent underestimation of the time and risks involved in a project. A manager sees challenges in a project as something that can be overcome by the team's high skill level, and downplay or ignore the risk of problems that are out of the team's control. It is for this reason that it does not matter if the project is broken down to the highest level of detail for cost prediction, all complex projects are at risk of encountering a multitude of problems that the manager could never foresee. Each problem has a low chance of occurring, but in combination the risk is much greater [@Lovallo2003]. @Lovallo2003 call the practise of analysing a project based on project details and its unique complexities the 'inside view'. Research has shown that if people are forced to make predictions about their skills after being exposed to an 'outside view', their predictions are significantly more accurate. This outside view can be applied to complex projects, and involves ignoring the details of the current project and instead analysing the outcomes of several projects of a similar description. This is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommend is to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model which is used to adjust the forecasted cost made by detailed analysis (inside view) [@Flyvbjerg2011]. The traditional way to think about a project is to focus on the project details. Even though gathering data from similar projects could significantly improve a cost prediction, it is rarely thought of by managers [@Flyvbjerg2011].

The implementation of reference class forecasting began in project management for the first time in 2004, was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011].

@@

Given this theory of reference class forecasting was first clearly published in 2003, the idea of predicting project costs from previous data has been the subect of research for many years before that and continues to this day. Although research results have often been promising, industry uptake of the idea is not yet established.

First an overview of the methods used will be presented followed by published theories for the lack of industry uptake.

@@

A substantial amount of research on project costprediction has been dedicated to the fields of software development and construction projects in particular. These analyses have generally used project information from 15 to 20 large projects. A single dataset may consist of projects from around the world, and each project a different client and project team. The mathematical methods used to predict project costs vary from statistical regressions to deep-learning neural networks. Most studies report neural networks are more accurate than regression

Previous research has been done on the prediction of final costs of a complex project with varying success. In the prediction of IT software development, formal estimation methods have existed for many years, however expert judgment is still the dominant method and available evidence does not suggest the models improve on expert judgment (Jorgensen and Shepperd, 2007). Construction industry research has used models ranging from neural networks to case based reasoning to regression to predict final building costs. However, often expert judgment was not mobilised, and the models were not comprehensive enough to be used in industry. Manager's experience and intuition is still the dominant method for construction cost calculation.

*** ^^^more research^^^ ***

Potential solution:
* stat analysis like reference class forecasting and building construction except analysis performed internally using the company's historical data on past projects.

case study:

This project undertakes a case study on a consulting company in the construction industry. Specifically, the

* in the case study (and is typical of smaller organisations??), the CRM is an untapped source of information storing
    * employee hours
    * client information/characteristics
    * invoicing
    * employee costs
    * employee charge out rates
    * project information
* currently the case study company uses a widely popular CRM, where they can output information but cannot perform modelling or analysis:
    * look up one job at a time
    * basic summaries of overall hours spent for the month

### 1.1.1 Limitations

* looking at a single case study company in a specific industry
* Managers commonly price jobs based on the budget for the entire project, of which the consultant is a minor part. An example of this is a consultant who typically charges 1% of the total project cost
    * this data has not been historically recorded. Can start being recorded and would be extremely useful
    * even better to have expected cost of entire project at start of project, as well as actual final cost of project.
* A fee is based off detailed information that is unique to each project, for example the size of a building in terms of storeys and square meterage. This detailed information is not available in the data, and is too complex and varied for an algorithm to digest. Therefore it is not reasonable for an algorithm to predict the fee of a project at this stage, but much more achievable to predict the profitability of a project (ie return per dollar) based on historical data. This profitability prediction relates to the profitability of the project if the fee is generated using the same method the company has used in its historical data.
* Although the aim is to predict the profitability of a project, the model aims to change managers' fee setting decisions if the algorithm predicts low profit or a loss based on historical data. 
* external factors forcing price down in some cases - competiton. clients

@@

```{r, echo=FALSE, include=FALSE}

read.csv(all8c,'C:/Users/n9232371/Documents/Consultbusiness/data/all8c.csv' )

vars<- names(all8c)
vars<- vars[-c("Business", "JD_Second", "code_client")]

>>>>

Project decision makers have struggled with forecasting project costs accurately for decades, across many industries. A study on large-scale infrastructure projects over the past seventy years revealed that cost forecasts consistently underestimated the cost of rail projects by 44.7%, the cost of bridge and tunnel projects by 33.8% and the cost of road projects by 20.4% (Flyvbjerg, 2006). A study across 1471 IT projects showed that 27% of projects ran over budget, and one in 6 of those projects were more than 200% over budget on average (Flyvbjerg and Budzier, 2011). These statistics cast a concerning glimpse at the financial story playing out behind the scenes of these projects which can cause job losses, ruin private businesses, and churn through government balancesheets. Complex long term projects are necessary to produce  railways, houses, websites, and movies. This research focuses on the financial risk that is transferred to consulting companies who are enlisted to complete portions of these complex projects. The risk is transferred by requiring a fixed flat fee from the consultant, meaning they commit to complete their role for a nominated price. In a similar way to infrastructure and IT projects, consultants are susceptible to underestimating the amount of their time that will be spent in their role, and must wear the cost of extra employee hours spent. For privately owned businesses, mitigating this risk would increase the chances of the business surviving as well as growing and succeeding.

The ramifications of unprofitable projects in a private business are multifold. Employee morale deteoriates and employees may view themselves as incompetent if they are responsible for the project's delivery. Employees also experience stress attempting to complete the project within a disappearing budget and may produce lower quality work in exchange for speed. This is unfortunate if in reality, the project fee was under estimated. Unprofitable projects also limit a business' opportunity to invest money into marketing or staff training, which would improve a business' standing. The financial risk taken on fixed flat fee projects may also discourage business' from taking different risks with better potential pay-offs, such as expanding the business into a new area. Logically, predicting fixed fee project costs more accurately could positively affect a business' growth opportunities and staff morale.

@@

The challenge for consultants to reduce the risk of a project going over budget, in terms of hours spent, is not easily solved. Complex projects are always different from each other, and similar work may take significantly different amounts of time. For example, an ad filmed in a remote location for chocolate will cost significantly more than an ad filmed on an exisitng set, even though the product structure is almost identical. Therefore, a fixed product price, which you could apply to products such as appliances, is inappropriate for film projects. One option is charging for consultants' time and the hiring of materials by the hour; a zero risk fee structure. However in many industries this is not competitive enough. Other consultancies may be willing to offer a fixed flat fee for the same work, reducing financial risk for the client, and luring the client to them. Another fee strategy may be to track worker hours, and stop work once the consulting fee has been depleted. Then, negotiate further fees, or variations, with the client and persuade them to accept. This also satisfactorily reduces consultant financial risk, and in some cases, depending on the nature of the work or uncontrollable changes, this method can be effective. However, if the project has been following a predictable process but the hours are nevertheless inflating beyond expectation (a common situation), consultants may prefer to wear losses or endure very marginable profits to preserve the relationship with their client. Consultant's may believe it is in their best interest to maintain a reliable, trustworthy reputation in their industry over creating financial friction. Clearly, there is no straightforward solution to reducing the financial risk taken on by consultant's who offer fixed fees in complex projects. In some indusries, fixed flat fee structures are expected and business' must conform to survive.

@@

Limited research is available on the cost estimation techniques performed by industry for complex projects, however two surveys were performed in 2000: one on construction projects, the other on IT projects. The construction cost estimation study by @Akintoye2000 surveyed 84 UK construction contractors, ranging from small to medium to large, about their estimating practises. They found that the main method for cost estimation was breaking the project into detailed parts and adding up the cost of each item. The next two most popular methods were 'comparison with similar projects based on documented facts', and 'comparison with similar projects based on personal experience'. These can all be classified as experience based models [@Akintoye2000]. A survey by Moores and Edwards in 1992 of 54 software developing companies found that detailed project planning tools were used by most companies as opposed to cost estimation tools, suggesting that projects were priced based on an analysis of a detailed breakdown of tasks within a project. In the same survey, 91% of software companies cited cost estimation as a problem. Although limited recent literature is available on industry cost estimation practises, historically it can be observed that for complex construction and software projects, cost predictions are chiefly made with a mixture of experience, project comparisons, and an analysis of the project details and tasks. It is also acknowledged that cost estimation is a problem and hypothesised that inexperience and lack of time are chief contributors [@Akintoye2000].

@@

@Lovallo2003 tackled the psychology behind the high failure rates of executives in predicting costs of projects such as manufacturing plant construction, mergers and acquisitions, large infrastructure and software development. Their theory stems from Kahneman's work on decision making that won him the nobel prize for economics in 2002. His research argues that humans' natural optimistic view of their own skills leads to consistent underestimation of the time and risks involved in a project. A manager optimistically sees challenges in a project as something that can be overcome by the team's high skill level, and downplay or ignore the risk of problems that are out of the team's control. It is for this reason that it does not matter if the project is broken down to the highest level of detail for cost prediction, all complex projects are at risk of encountering a multitude of problems that the manager could never foresee. Each problem has a low chance of occurring, but in combination the risk is much greater [@Lovallo2003]. @Lovallo2003 call the practise of analysing a project based on project details and its unique complexities the 'inside view'. Research has shown that if people are forced to make predictions about their skills after being exposed to an 'outside view', their predictions are significantly more accurate. This outside view can be applied to complex projects, and involves ignoring the details of the current project and instead analysing the outcomes of several projects of a similar description. Mobilising the outside view in this way is called reference class forecasting [@Lovallo2003]. One method @Lovallo2003 recommend is to obtain correlation statistics from past similar projects - the correlation between the forecast cost and the actual cost. The correlation for the current project can then be estimated via a statistical model which is used to adjust the forecasted cost made by detailed analysis (the inside view) [@Flyvbjerg2011]. The traditional way to think about a project is to focus on the project details. Even though gathering data from similar projects could significantly improve a cost prediction, it is rarely thought of by managers [@Flyvbjerg2011].

The implementation of reference class forecasting began in project management for the first time in 2004, was endorsed by the American Planning Association in 2005, and is now used in some governments and private companies in Europe, South Africa, and Australia. An example of the type of output from a reference class statistical model is a plot showing the relationship between the acceptable chance of cost overrun and the required uplift to the original forecasted cost [@Flyvbjerg2011].

@@

The theory of reference class forecasting was first clearly published in 2003, however the idea of predicting project costs from previous data has been the subject of research for many years before that and continues to this day. Although research results have often been promising, industry uptake of the idea is not yet established.

### 1.1.1 Overview of Past Reference Class Forecasting Methods and relevant Industry Uptake

In the following section, an overview of the historical attempts to integrate reference class forecasting will be presented followed by published theories for the lack of industry uptake.

@@

#### 1.1.1.1 Existing Models

A substantial amount of research on project cost prediction has been dedicated to the fields of software development and construction projects in particular. These analyses have generally used project information from 15 to 20 large projects. A single dataset may consist of projects from around the world, and each project a different client and project team. The mathematical methods used to predict project costs vary from statistical regressions to deep-learning neural networks [@Love2005][@Kim2004]. Most studies report neural networks are more accurate than regression, and generally the studies report positive predictive powers of their developed models.

A McKinsey study outlined a case in the film industry where a motion picture company used reference class forecasting of movie project success (cases were weighted by similarity) to decide which movies should be heavily promoted. This model improved forecasts on financial return by 135% relative to single project analogies [@Flyvbjerg2014].

Statistical research projects which trial reference forecasting have shown promising results in IT, construction, and film projects. However, there is little evidence that the methods have successfully transitioned into industry.

#### 1.1.1.2 Industry Uptake of Existing Models and Current Practises

Surveys on current industry practise for cost estimation are scarce, however, a survey of the construction industry in the UK in 2000 found that a manager's experience and intuition is still the dominant method for construction cost calculation [@Akintoye2000]. A study on the IT industry by @Jorgensen2007 cited that the parametric models were not comprehensive enough to be used in industry and also found that expert judgment was not mobilised enough in the process of developing these tools.

Although reference class models have not gained traction in either the software or construction industry (where most of the research to date has focussed), it is worth outlining their current practises in cost estimation and the reasons behind these.

In the construction industry, infrastructure and building projects follow a traditional and well established process, where contractual norms have developed over many decades [@Badenfelt2011]. The industry is so large, broad and well established that it has inertia in the way that projects are estimated, and contractually bound. The contractor (builder) first reviews the drawings and if the plans are not detailed enough, they offer a temporary budget price. Once they feel that the drawings are sufficiently detailed, they will offer a fixed price [@Badenfelt2011]. However, as the job progresses, if any work is added that was not in the detailed drawings, they maintain the right to charge a variation [@Badenfelt2011]. This method is clear and variations are easily justifiable to the client, however cost, labour and time estimates are still regularly under calculated, even from detailed drawings. And for large infrastructure projects, small percentage miscalculations can be exceedingly expensive.

Contractor's practise of pricing from detailed drawings and charging for variations does not translate to other consultants in the construction industry, such as engineers and architects. Consultants must devlop the detailed plans for the contractors, and must therefore offer fixed price contracts based on drawings with little detail. It is much harder for a consultant to charge a variation for a portion of their time spent on something that was not included in the original simple plans. This is despite the real and common instance of budget blow outs due to overspent time [@Harris1999]. One method to provide opportunity for a variation is stating a minimum percentage fee of the final building cost, which can overrule the original fixed price contract. If a consultant has spent time beyond their fixed price budget,this may account for some or in the best case, all, of their overspent time.

Construction projects generally are completed more on time and on budget when compared to IT projects as the greatest cost is the project is building materials [@Badenfelt2011]. Unlike time, building materials can be accurately calculated from detailed design drawings. Again, this omits data from consultants who are part of the building industry but charge based on their time only.

In contrast to construction, IT projects are a relatively new practise where alternative contractural arangements have been trialled. Through experimentation with contracts, the Agile method was developed. In summary, the method treats both cost and time as fixed quantities for a project, and if any change or additional work can only be accommodated if another, less important requirement is excluded [@Badenfelt2011]. Furthermore, the project outcomes are continually adjusted ad revised based on frequent progress review points that assesses unexpected problems. This agile attitude has become ingrained in many IT practises and this flexible upfront arrangement with clients often forges better long term relationships [Cockburn2001].

It is not completely clear why industry uptake of cost estimation tools using the outside view has been weak, and it may be the case that in modern industries such as in the IT industry, problems associated with cost estimation can be innovatively designed out of the contract. Possible reasons that parametric reference class forecasting models developed by researchers have not been incorporated into industry practise include:

* lack of appropriate data
    * if a project is unique, such as a national infrastructure project, it many not feel appropriate to gather sufficient data points from other countries and time periods.
* lack of understanding from industry decision makers - the prediction process was not explained well in marketing of the tools
* models were not developed in collaboration with industry
* does not feel intuitive price from an 'outside view' of a project which disregards small project details
* insufficient time allocated to pricing that makes change burdensome

In fact, a study by @Moores1992 concluded that a lack of framework to support the outside view model use and a failure in marketing were the two most common explanations for lack of uptake of cost estimating tools which use statistical techniques in predicting project costs. This is unfortunate as almost all research indicates an outside view analysis of a job substantially improves cost estimation accuracy for fixed price projects.

### 1.1.2 Case Study

This research will explore the potential for reference class forecasting to improve cost estimation in the context of an engineering consulting company in the construction industry. In this field, labour is the chief cost and the long traditional history of this industry mandates fixed price projects as the norm. In recent years, the size of losses from unprofitable projects in the case study could equal up to 25-30% of the profits from profitable projects. This demonstrates substantial room for improvement and the possiblity of multitude positive flow on effects.

Cost estimation is currently performed by first carefully reviewing preliminary drawings for a project where time costs will be calculated using a manager's personal experience with similar projects. This will often be cross-checked with a value based off a percentage of the estimated final cost of construction (generally the client budget). No formal mathematical method of comparison to similar projects is performed. This is the case for a number of possible reasons:

* lack of time to interrogate the CRM
* clunky data availability offered by the CRM
* lack of awareness of projects performed by other managers that may have been similar.

Reference class forecasting in this study differs from previous research, in that statistical analysis will be performed on internal past company project data that has been progressively recorded in the company's Customer Relationship Management (CRM) software which was purchased. This presents an opportunity to evaluate reference class forecasting in a unique situation, where the 'reference' projects are internal and over two thousand cases are available.

The CRM data currently stands as an untapped source of information within the case study organisation. It stores a rich variety of data including:

* employee timesheet hours with dates
* other project costs (taxis, printing)
* client information/characteristics
* client identification code
* invoiced amounts for each project and dates
* employee costs
* employee charge out rates
* project description

A statistical model predicting a measure of project financial success will be developed using the available project variables and trialling numerous machine learning algorithms and statistical methods. The company uses well established CRM software which readily provides simple output statistics such as project time-cost vs invoiced amount summaries. However, the analytical capabilities were limited to:

* simple scatter plots and bar charts of the raw data
* summaries such as overall hours spent vs invoiced amount for each project

The company has a wealth of data but limited means to extract insight. This study will exploit the CRM data by performing more sophisticated statistical analysis with the intention of building a predictive model for cost estimation. The benefits of this study in comparison to previous cases are that thousands of past data points are available, when previous studies used on average 15-20 cases and at most 170. This improves the potential for accurate predictions. If successful, the model also has higher potential for managerial uptake as the algorithm will be directly built and trained on company data. Managers can interrogate 'similar' data points and relate to the actual projects or speak to a colleague who was involved. This has the potential to give the 'outside view' stronger influence on the final decision.

### 1.1.3 Case Study Limitations

The case study provides a framework to test the value of a consulting company's CRM data in assisting project cost estimates. However, it is important to be aware of limitations when reviewing outcomes and framing the problem. The most obvious limitation is that we are disection a single company in a specific industry, which limits the ability to state conclusions could apply to a broad range of cases. This can be overcome with future studies.

The data set from this case study does not include information that would improve project cost predictions. For example, managers commonly price jobs based on the budget for the entire project, of which the consultant is a minor part. The manager might typically moderate his fee by say 1% of the entire project cost. The value of the entire project has typically not been recorded although there is a space in the project software form for this value. It would be even more beneficial for managers to record the expected total project cost at the beginning of a project versus the conclusion of a completed construction project. Another piece of data that was not recorded was external factors influencing the fixed price quote including competition from similar firms that the client played against the case study company.

In addition to this data that was not typically entered by workers, the dataset also excludes detailed information about the projects which are crucial to the final pricing. This includes details such as preliminary drawings describing the project (square meterage, number of storeys), and project summaries provided by the client to enable a quote. This kind of detailed information will never be recorded in a CRM and in many ways contradicts the purpose of the 'outside view' which intends to present similar projects without being over-influenced by the finer details of a project. Given the lack of detailed information, it is not reasonable for an algorithm to predict the specifc fee of a project. It is much more effective to attempt predict the profitability of a project (ie return per dollar) based on the historical data. This profitability prediction relates to the profitability of the project if the fee was generated using the established method the company used in its historical cases.

Finally, CRM data is entered by employees manually as time passes, jobs are invoiced, and projects are finally completed. Because the data is input by humans, it is susceptible to input error. Many of errors were detectable during data cleaning, however it is possible that undetectable human errors exist in the data.

Although this case study presents an opportunity to test the predictive power held within the organisations' existing CRM data, limitations exist which must be understood and built into the framework for predictions. These include missing data that could have been entered for each project, but historically has not been required, missing data that is very detailed and will never be available, and the potential for a small degree of human error to be embedded within the data.

### 1.1.4 Research Motivation Summary

Fixed price cost estimation for complex projects is a difficult task and when performed incorrectly, can have devastating affects on businesses and governments. Many theories have delved into human nature and our tendency to optimistically assess our capabilities with respect to a given task. In this case, the optimism often results in negative business consequences affecting employee morale and the capacity of a company to flourish and grow. Over the past few decades, significant research has been dedicated to creating predictive models that present the 'outside view' of a project, by statistically comparing a new project to numerous similar projects and their characteristics. It has been shown that these models improve the cost estimation, however industry uptake has not been successful. It has proven to be a challenging task to persuade managers to use this so called reference class models in practise and remove themselves from additions based on intricate details of the completed project. This may be due to a number of reasons including lack of understanding, time, and lack of collaboration with industry in development of these models.

@@

```{r, echo=FALSE, include=FALSE}
library('magrittr',lib='C:/Progra~1/R/R-3.2.2/library')

all8c<- read.csv('C:/Users/n9232371/Documents/Consultbusiness/data/all8c.csv' )

vars<- names(all8c)
vars<- vars[!vars %in% c("Business", "JD_Second", "code_client")]

### 2015-11-09 financial.rmd
save early nov

```{r_loadpackages, echo=FALSE, include=FALSE}
library("plyr",lib = 'C:/Progra~1/R/R-3.2.1/library')
library("dplyr",lib = 'C:/Progra~1/R/R-3.2.1/library')
library('magrittr',lib='C:/Progra~1/R/R-3.2.1/library')
library('reshape2',lib='C:/Progra~1/R/R-3.2.1/library')
library('ggplot2',lib='C:/Progra~1/R/R-3.2.1/library')
library('gbm', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('caret', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('ROCR',lib='C:/Progra~1/R/R-3.2.2/library')

@@

#without algorithm, balance would be $1,475,642
sum(prof$balance)

# without any losing jobs, balance would be $2,516,246
prof %>% filter(balance > 0) %>% sum

# without any profitable jobs, balance would be -$1,040,322
prof %>% filter(balance < 0) %>% sum

@@

>>>>

```{r_loadpackages, echo=FALSE, include=FALSE}
library("plyr",lib = 'C:/Progra~1/R/R-3.2.1/library')
library("dplyr",lib = 'C:/Progra~1/R/R-3.2.1/library')
library('magrittr',lib='C:/Progra~1/R/R-3.2.1/library')
library('reshape2',lib='C:/Progra~1/R/R-3.2.1/library')
library('ggplot2',lib='C:/Progra~1/R/R-3.2.1/library')
library('gbm', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('randomForest', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('caret', lib = 'C:/Progra~1/R/R-3.2.2/library')
library('pROC',lib='C:/Progra~1/R/R-3.2.2/library')

@@

#without algorithm, balance would be $1,475,642
orig.bal<- sum(prof$balance)

# without any losing jobs, balance would be $2,516,246
orig.prof<- prof %>% filter(balance > 0) %>% sum

# without any profitable jobs, balance would be -$1,040,322
orig.loss<- prof %>% filter(balance < 0) %>% sum

@@

Write function for making 10 boosted trees and checking the predictions/savings for each!

all10mice<- read.csv('C:/Users/n9232371/Documents/Consultbusiness/data/all10mice.csv')[,-1]

profit.cv<- function(df.string = "all10mice", method = "log", seed = 300, k.folds = 8){

        results = data.frame('orig.bal' = rep(0, k.folds), 'orig.prof' = rep(0, k.folds), 
                             'orig.loss' = rep(0, k.folds),
                             'prof.dismis' = rep(0, k.folds), 'loss.dismis' = rep(0, k.folds), 
                             'algor.prof' = rep(0, k.folds),
                             'improv' = rep(0, k.folds), 'thresh' = rep(0, k.folds),
                             'auc' = rep(0, k.folds))
        colours <- c("red", "yellow", "green", "blue", "purple")

        df = get(df.string)
        set.seed(seed)

        # create f.rpdol
        df$f.rpdol <- as.factor(df$b.rpdol)
        levels(df$f.rpdol)[levels(df$f.rpdol) == "0"] <-"profit"
        levels(df$f.rpdol)[levels(df$f.rpdol) == "1"] <-"loss"

        #set up 10 fold cv using caret function
        folds <- createFolds(df$f.rpdol, k = k.folds, list = T, returnTrain = F)
        #now have folds[[1]] through folds[[10]] list of rows to exclude

        if(method %in% 'boost') {
                        formula = "b.rpdol ~ Discipline + pc.pro + b.timespan.cbrt + no.users + inv.mlsto.log + client.totinv.log + Business + majority.pos + pc.majpos.log"
        } else {
                formula <-"f.rpdol ~ Discipline + pc.pro + b.timespan.cbrt + no.users + inv.mlsto.log + client.totinv.log + Business + majority.pos + pc.majpos.log + JD.Second + Billing.Type"
        }

        for (j in 1:length(folds)) {

                        #turn predictors and response into 'ordered' variables for roc() to work
                        test<- df[folds[[j]],]
                        levels(test$f.rpdol) <- c('profit', 'loss')
                        test$f.rpdol <- ordered(test$f.rpdol)

                        if(method %in% 'rf'){
                                fit <- randomForest(as.formula(formula), data = df[-folds[[j]],], 
                                            mtry = 5, ntree = 2000)
                                pred <- predict(fit, test, type = "prob")
                                pred <- pred[,2]

                        } else {
                                if(method %in% 'log'){
                                        fit<- glm(as.formula(formula), 
                                                  data = df[-folds[[j]],], family = binomial())
                                        pred <- predict(fit, test,
                                                        type = "response")
                                } else {
                                        fit <- gbm(as.formula(formula), data = df[-folds[[j]],],
                                                   distribution = "bernoulli", n.trees = 10000,
                                                   shrinkage = 0.001, interaction.depth = 4,
                                                   n.minobsinnode = 20)

                                                pred <-predict(fit, test, 
                                                               n.tree = 10000, type = "response")
                                                }
                        }

                        #pROC
                        pred.p <- roc(test$f.rpdol, pred)

                        results$auc[j] <- auc(pred.p)

                        ##
                        #create full test df
                        test$pred<- pred
                        test.pred <- merge(test, all7d %>% select(mlsto, balance.mlsto), by = "mlsto")

                        #create vector of profit points

                        threshold = seq(0,1, by = .05)
                        fin.prof = rep(0, length(threshold))

                        for (i in 1:length(threshold)) {
                                prof <- data.frame("balance" = test.pred$balance.mlsto,
                                "pred" = test.pred$pred,
                                "ans" = test.pred$b.rpdol)

                                #pred = probabilities
                                prof$conf <- rep(0, nrow(prof))
                                prof$conf <- ifelse(prof$pred >= threshold[i] & prof$ans == 1, 'TP', prof$conf)
                                prof$conf <- ifelse(prof$pred >= threshold[i] & prof$ans == 0, 'FP', prof$conf)
                                prof$profit <- rep(0, nrow(prof))
                                prof$profit <- ifelse(prof$conf %in% "TP",-prof$balance, prof$profit)
                                prof$profit <- ifelse(prof$conf %in% "FP",-prof$balance, prof$profit)

                        print(
                        ggplot(prof.plot, aes(x=threshold, y=final.profit)) + geom_point() + geom_line()
                        )

                        best.thresh = arrange(prof.plot, -final.profit)[1,1]
                        results$thresh[j] = best.thresh

                        #without algorithm, balance would be $1,475,642
                        results$orig.bal[j] = sum(prof$balance)
                        # without any losing jobs, balance would be $2,516,246
                        results$orig.prof[j] = prof %>% filter(balance > 0) %>% sum
                        # without any profitable jobs, balance would be -$1,040,322
                        results$orig.loss[j] = prof %>% filter(balance < 0) %>% sum
                        #how much real profit would you be dismissing? $82631
                        temp<- prof %>% filter(pred >= best.thresh & balance > 0)
                        results$prof.dismis[j] = ifelse(nrow(temp)>0, temp %>% select(balance) %>% sum,
                                                        0)
                        #how much real loss would you be avoiding? $566,481.6
                        results$loss.dismis[j] = prof %>% filter(pred >= best.thresh & balance < 0) %>% 
                                select(balance) %>% sum

                        #print j value
                        cat(j, " ")

        # total benefit = $483,850
        results = transform(results, algor.prof = -(loss.dismis + prof.dismis))
        results = transform(results, improv = algor.prof/orig.bal)

        return(results)

results.1 <- profit.cv(df.string = "all10mice", method = "log", seed = 400, k.folds = 8)
results.2 <- profit.cv(df.string = "all10mice", method = "boost", seed = 400, k.folds = 8)
results.3 <- profit.cv(df.string = "all10mice", method = "rf", seed = 300, k.folds = 3)

### 2015-11-09 rf_log_naiive_auc.rmd
save early nov

                #set up 10 fold cv using caret function
                folds <- createFolds(df$f.rpdol, k = 10, list = T, returnTrain = F)
                #now have folds[[1]] through folds[[5]] list of rows to exclude

>>>>

                #set up 10 fold cv using caret function
                folds <- createFolds(df$f.rpdol, k = 10, list = T, returnTrain = F)
                #now have folds[[1]] through folds[[10]] list of rows to exclude
